Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2022-02-18 Manuscript/Manuscript.tex
as

\def\Authors{
  Maxwel C Oliveira\,\textsuperscript{1},
  Amit J Jhala\,\textsuperscript{2},
  Mark Bernards\,\textsuperscript{3},
  Chris Proctor\,\textsuperscript{2},
  Strahinja Stepanovic\,\textsuperscript{2},
  Rodrigo Werle\,\textsuperscript{1*}}

@@

  \textsuperscript{3} Department of Agronomy, Western Illinois
University,  Macomb,  Illinois,  United States
  }

@@

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
ranked as one of the most economically detrimental weed species to
cropping systems in the United States (Van Wychen, 2020). Unmanaged
Palmer amaranth plants compete for water, light, and nutrients, which
can drastically impact crop yields (Berger et al., 2015). For example,
Palmer amaranth has been documented to reduce up to 91\%, 68\%, and 54\%
corn (Massinga et al., 2001), soybean (Klingaman and Oliver, 1994), and
cotton (Morgan et al., 2001) yields, respectively. Moreover, Palmer
amaranth has shown a remarkable capacity to evolve resistance to
herbicides. To date, Palmer amaranth has evolved resistance to eight
herbicide sites of action (Heap, 2021), increasing the weed management
complexity (Lindsay et al., 2017). Thus, Palmer amaranth poses an
economical and ecological risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selections of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecious nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth's small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), seed mixes (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth emergence is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Also, Palmer amaranth has shown to grow and
develop under reduced soil water content. In continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds
plant\textsuperscript{-1} (Chahal et al., 2018). Seeds from Palmer
amaranth growing with limited water conditions were heavier, less
dormant, and prompt for germination (Matzrafi et al., 2021). Growing
conditions and management practices also influence Palmer amaranth sex
dimorphism and flowering pattern (Korres et al., 2017; Rumpa et al.,
2019). Therefore, Palmer amaranth has shown plasticity to evolve and
fast adapt under the current agroecosystem conditions. Future scenarios
suggest that global temperature warming can impact agriculture, and
promote niches for Palmer amaranth invasion/adaptation into new
environments. Agronomic crops in Australia and Sub-Sahara Africa are
currently at greatest risk to Palmer invasion as climate warms (Kistner
and Hatfield, 2018). Temperature is a key factor limiting Palmer
amaranth expansion to cooler geographies (Briscoe Runquist et al.,
2019); however, under future climate change Palmer amaranth is likely to
expand northward into Canada and Northern Europe (Kistner and Hatfield,
2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause yield loss to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn comprise most
of US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021), which is considered the
most troublesome species in the US Midwest (Tranel et al., 2011).
Invasion and successful eradication of Palmer amaranth is documented in
Minnesota (Yu et al., 2021). Palmer amaranth infestations have not been
detected in Canada; however, Palmer amaranth seeds were detected in
sweet potato slips in the country (Page et al., 2021). Palmer amaranth
is still not as well adapted and established to Northern as compared to
Southern North America. Therefore, Palmer amaranth range of expansion
into new habitats can increase. It seems certain the need to manage new
Palmer amaranth infestations in agronomic crops throughout northern
United States is eminent in the near future. Strategies on Palmer
amaranth management should encompass the agroecosystem level but not
focus only on attempts to eradicate the weed. Most tactics to manage
Palmer amaranth are based on technological solutions (Scott, 2011),
which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built
around minimizing the species ability to adapt, grow and develop
successfully in agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in northern states are limited due to the plant classification
as noxious weed species (Yu et al., 2021). Nonetheless, the continuous
Palmer amaranth dispersal and potential establishment across the
northern United States is concerning and warrants investigations on
species morphology in northern environments. Understanding Palmer
amaranth biology and morphology under different agroecosystems can
enhance our knowledge on species adaptation and management practices. It
can also aid in designing proactive and ecological tactics to limit the
species range expansion, reduce its negative impact on agroecossystems,
and to develop resilient and sustainable farming systems (MacLaren et
al., 2020). Therefore, the objective of this study was to investigate
the flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing under corn, soybean and bareground across five
locations in the United States Midwest.

@@

A Palmer amaranth accession (Kei3) from Perkins County, Nebraska with no
glyphosate resistance according to Oliveira et al. (2021) was selected
for this study. Three weeks prior to the establishment of each cohort,
seeds were planted in plastic trays containing potting-mix. Emerged
seedlings (1 cm) were transplanted into 200 cm\textsuperscript{-3}
plastic pots (a plant pot\textsuperscript{-1}). Palmer amaranth
seedlings were supplied with adequate water and kept under greenhouse
conditions at the University of Wisconsin-Madison, University of
Nebraska-Lincoln, and Western Illinois University; and kept outdoors at
the Perkins extension office in Grant, NE until the 2-3 leaf stage (5 to
8 cm height) when they were transported to the field.

@@

Fields were conventionally tilled prior to crop planting. Corn and
soybean were planted in 76-cm row spacing (Table 1). Monthly mean air
temperature and total precipitation were obtained using Daymet weather
data from June through September across the five locations in 2018 and
2019 (Correndo et al., 2021) (Figure 1)

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each field
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplant timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting to assure seedling survival.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant
gender was identified (male or female), plant height was measured from
soil surface to the top of plant. Also, aboveground plant biomass was
harvested near soil surface and oven dried at 65 C until reaching
constant weight before weigh (g plant\textsuperscript{-1}) was recorded.

Plants had to be harvested at flowering because Palmer amaranth is
neither endemic in Wisconsin nor in Illinois. In our study, all
locations followed the methodology of plant harvest at flowering
initiation, except in Grant, NE. In this location, all Palmer amaranth
plants were harvest at once on July 06, 2018 and 2019 (first cohort),
and on August 17, 2018 and on July 31, 2019 (second cohort).

@@

Analyses of Palmer amaranth height and biomass were performed with a
linear mixed model using \emph{lmer} function from ``lme4'' package
(Bates et al., 2015). Plant height and biomass were log transformed to
meet model assumption of normality. In the model, crop (bareground,
corn, soybean) and cohort time (first and second) were the fixed effects
and year nested with location the random effects. Analysis of variance
at \(\alpha\) 0.05 was performed with \emph{anova} function from ``car''
package (Fox and Weisberg, 2018). Marginal means and compact letter
display were estimated with \emph{emmeans} and \emph{cld} from packages
``emmeans'' (Lenth et al., 2021) and ``multcomp'' (Hothorn et al.,
2008), respectively.

The Palmer amaranth flowering timing was estimated as cumulative
flowering across all locations, except Grant, NE. Palmer amaranth
cumulative flowering estimation was determined using an asymmetrical
three parameter log logistic Weibull model of the drc package (Ritz et
al., 2015).

@@

In this model, \emph{Y} is the Palmer amaranth cumulative flowering,
\emph{d} is the upper limit (set to 100), and \emph{e} is the inflection
point, and \emph{x} day of year (doy).

The doy for 10, 50, and 90\% Palmer amaranth cumulative flowering were
determined using the \emph{ED} function of drc package. Also, the 10,
50, and 90\% Palmer amaranth cumulative flowering were compared among
crops and cohorts using the \emph{EDcomp} function of drc package. The
EDcomp function compares the ratio of cumulative flowering using
t-statistics, where P-value \textless{} 0.05 indicates that we fail to
reject the null hypothesis.

A binary logistic regression was fitted to Palmer amaranth gender.
Binary logistic regression is used for predicting binary classes
(Bangdiwala, 2018), such as the probability of a plant being female in a
dioecious species. Prior to the analysis, missing values were removed
from the dataset. Also, data from Grant was not used in this analysis
due to the uniform plant harvesting at that location. The complete
dataset was splitted into 80\% train and 20\% test data. The 80\% train
is used for the model training and the 20\% test is used for checking
model performance on unseen dataset. With 80\% dataset, a generalized
linear model (base R \emph{glm} function) was fitted to binary response
variable, the probability of being female (0 to male and 1 to female).
The independent variables were day of year harvest, height, weight, and
crop (without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function from ``rcompanion'' package (Mangiafico,
2021). The marginal effects computation was performed with Average
Marginal Effects (AMEs) at every observed value of x and average across
the results (Leeper, 2017) using \emph{margins} function from
``margins'' package (Leeper et al., 2021). The 20\% test data was
predicted using the \emph{predict} function with a cutoff estimation for
male or female using \emph{performance} function from ROCR package (Sing
et al., 2005). The model quality prediction from the classification
algorithm was measured with precision (\emph{precision} function),
recall (\emph{recall} function) and F1-score (\emph{f\_meas} function)
using the ``yardstick'' package (Kuhn et al., 2021). The precision
determines the accuracy of positive predictions (female plants), recall
determines the fraction of positives that were correctly identified, and
F1-score is a weighted harmonic mean of precision and recall with the
best score of 1 and the worst score of 0 (Raoniar, 2021). F1-score
conveys the balance between the precision and the recall (Yacouby and
Axman, 2020). The area under the receiver operating curve (AUC-ROC) was
also estimated with performance function using the true positive and
false positive rates. Higher the AUC, better the model is at
distinguishing between female and male Palmer amaranth.

@@

Palmer amaranth plants accumulated more biomass when growing in
bareground compared to plants growing in soybean and corn (figure 3A).
Palmer amaranth plants in the first cohort produced 75.5, 28.3, and 16.3
g plant\textsuperscript{-1} in bareground, soybean and corn,
respectively. Plants from the second cohort produced 62.6 g
plant\textsuperscript{-1} in bareground, followed by 6.3 g
plant\textsuperscript{-1} in soybean, and 1.4 g
plant\textsuperscript{-1}.

Palmer amaranth height was more uniform across cohort timings, except
when growing in corn (figure 3B). Palmer amaranth plants from the first
cohort were on average 69.2 cm tall in bareground, which was not
different from the 70.7 cm tall plants from the second cohort timing (P
= 0.74). In addition, no difference in Palmer amaranth height (69.3 cm)
was detected from first cohort plants in soybean to first and second
cohort plants in bareground (P \textgreater{} 0.75). Palmer amaranth
plants from the second cohort were nearly 10 cm lower compared to the
first cohort in soybeans (P = 0.04). The tallest (first cohort) and
smallest (second cohort) Palmer amaranth plants were found in corn.
Palmer amaranth reached 85.2 and 38.2 cm tall, respectively.

@@

Palmer amaranth plants from the first cohort growing in corn resulted in
a longer flowering window compared to plants growing in bareground and
soybean (Figure 4A). The 10\% cumulative Palmer amaranth flowering in
soybean, bareground and corn occurred at the end of June. Palmer
amaranth reached 10\% flowering in soybean, bareground and corn at doy
180, 180.9 and 181.7, respectively. The 50\% Palmer amaranth cumulative
flowering occurred in July. Palmer amaranth reached 50\% flowering in
bareground, soybean and corn at doy 193.4, 194.8, and 206.6,
respectively. Similar trend was observed at 90\% Palmer amaranth
cumulative flowering. Palmer amaranth reached 90\% flowering at doy
252.6 in corn (early September), which was 38 and 32 days after reaching
90\% flowering in bareground and soybean, respectively.

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% less females plants (Table 2). Moreover, increasing a
unit doy increases the probability of having a female plant by 0.4\%
(Table 2 and Figure4A). Similar trend is observed for weight (Figure 4B)
as well as height (Figure 4C), whereas the probability of being female
increase by 0.2\% and 0.1\% when a unit of weight (g) and height (cm)
increases, respectively.

\caption{\label{tab:unnamed-chunk-3}Average marginal effects of Palmer amaranth gender logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

weight & 0.001 & 0.000 & 0.000 & 0.001 & 2.179 & 0.029\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} Average Marginal Effects. \textsuperscript{b} Standard Error.}\\
\end{tabular}
\end{table}

@@

Our study showed that Palmer amaranth biomass, height, flowering window
and gender varied within crops and cohort timings. In general, first
cohort of Palmer amaranth plants were heavier and taller when compared
to the second cohort. At first cohort, resources (e.g., soil nutrients)
and conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape. The Palmer amaranth competition (e.g., light)
strategy was to mimic the crop growth and development (Figure 6). These
results suggests that Palmer amaranth can quickly evolve life-history
traits to adapt to cropping systems and cultural practices, which was
also shown in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight Palmer amaranth as a threat to field crops
as breeding more competitive crop varieties is likely to select more
competitive weed biotypes (Bravo et al., 2017).

@@

@@

>>>>

\def\Authors{
  Maxwel C. Oliveira\,\textsuperscript{1},
  Amit J. Jhala\,\textsuperscript{2},
  Mark L. Bernards\,\textsuperscript{3},
  Chris Proctor\,\textsuperscript{2},
  Strahinja Stepanovic\,\textsuperscript{2},
  Rodrigo Werle\,\textsuperscript{1*}}

@@

  \textsuperscript{3} School of Agriculture, Western Illinois
University,  Macomb,  Illinois,  United States
  }

@@

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
ranked as one of the most economically detrimental weed species to
cropping systems in the United States (Van Wychen, 2020). Unmanaged
Palmer amaranth plants compete for water, light, and nutrients, which
can drastically impact crop yields (Berger et al., 2015). For example,
Palmer amaranth has been documented to reduce up to 91\%, 68\%, and 54\%
corn (Massinga et al., 2001), soybean (Klingaman and Oliver, 1994), and
cotton (Morgan et al., 2001) yields, respectively. Moreover, Palmer
amaranth has shown a remarkable capacity to evolve resistance to
herbicides. To date, Palmer amaranth has evolved resistance to eight
herbicide sites of action (Heap, 2021), increasing the weed management
complexity (Lindsay et al., 2017) and posing an economical and
ecological risk to row-crop agriculture.

Palmer amaranth as a problem weed is a function of both inherent
adaptations and selected management practices. Palmer amaranth is a fast
growing summer annual forb indigenous to the Sonoran Desert (Sauer,
1957). It became a serious problem weed in US agriculture in the 1990s
(Ward et al., 2013). Palmer amaranth weediness is likely a result of
human-assisted selection combined with plant biology. Farm
mechanization, adoption of conservation agriculture (e.g., no-till), and
intensive use of herbicides for weed management are the main
human-mediated selections of Palmer amaranth in cropping systems (Ward
et al., 2013). Palmer amaranth is a prolific seed producer with a C4
photosynthetic apparatus (Wang et al., 1992). With dioecious nature,
Palmer amaranth male and female plants are obligate outcrossers,
increasing the chances of exchanging adaptive traits among plants (Jhala
et al., 2021; Oliveira et al., 2018). Also, Palmer amaranth's small
seeds (e.g, 1 mm) tend to thrive in no-tillage systems (Price et al.,
2011), and spread across locations through farm equipment (Sauer, 1972),
seed mixes (Hartzler and Anderson, 2016), wildlife (Farmer et al.,
2017), etc., making it one of the most successful examples of weed
adaptation to current cropping systems.

Palmer amaranth's plasticity allows it to respond successfully to
environmental changes. Palmer amaranth demonstrates a high degree of
plasticity to light, temperature, water availability, and human
management (Jha et al., 2010). Palmer amaranth has an extended
germination period of up to \_\_\_\_\_ (Ward et al., 2013). Germination
of Palmer amaranth was triggered by 18 C soil temperature at 5 cm depth
(Keeley et al., 1987), and optimal germination and biomass production
occurred at 35/30 C day and night temperatures (Guo and Al-Khatib,
2003). Palmer amaranth emergence can be influenced by tillage and the
use of preemergence herbicides (Chahal et al., 2021), and may result in
weed germination shifts within a population, as documented in
\emph{Bassia scoparia} (Sbatella and Wilson, 2010). In an experiment
where Palmer amaranth was subjected to continuous water stress, it
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019).

Palmer amaranth's current global range includes agronomic cropland in
Italy (Milani et al., 2021), soybean producing regions of Brazil and
Argentina (Larran et al., 2017; Küpper et al., 2017), and the Cotton
Belt (Garetson et al., 2019; Bagavathiannan and Norsworthy, 2016) and
southern Great Plains of the United States (Kumar et al. (2020), Crespo
et al. (2016){]}. However, its range appears to be expanding steadily,
and in some cases rapidly, in the U.S. For example, in states with large
Palmer amaranth infestations its range is gradually pushing north as it
displaces other weeds in agronomic row crops (R Werle, personal
observation), aided by wildlife (Farmer et al., 2017) and agricultural
machinery. But there are other examples where it has suddenly become a
problem weed hundreds of miles away from known infestations. New
infestations in Michigan, Indiana and North Dakota appear to have
resulted from Palmer amaranth contaminated livestock feed. New
infestations in Iowa and Minnesota were associated with planting
contaminated pollinator habitat seed (Yu et al., 2021). Palmer amaranth
infestations have not been detected in Canada, but Palmer amaranth seeds
were found in sweet potato slips imported into the country (Page et al.,
2021). Global warming will create new opportunities for Palmer amaranth
invasion. Although agronomic crops in warm environments like Australia
and Sub-Saharan Africa are currently at greatest risk for Palmer
amaranth invasion, warming temperatures will reduce barriers that may
have limited Palmer amaranths spread into cooler climates like Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

There are many areas in the U.S. Corn Belt (North Central States) where
Palmer amaranth is not yet established, and its potential adaptability
is untested. Nonetheless, the rapid expansion of Palmer amaranth across
the Northern U.S. is concerning and warrants investigations on its
adaptability. We do know that Palmer amaranth caused yield loss in
Illinois soybean fields (Davis et al., 2015), but an Iowa study showed
that Palmer amaranth was not as well adapted as waterhemp
(\emph{Amaranthus tuberculatus}) to conditions in that geography (Baker,
2021). We also know that Palmer amaranth plants that establish shortly
after row-crop planting have a much greater impact on crop yield than
plants that emerge after the crop has produced several leaves (MacRae et
al., 2013). Understanding Palmer amaranth morphology and development
under different agroecosystems and across a wide geography can enhance
our knowledge of its adaptability, and may also aid in designing
effective tactics to limit is range expansion and minimize its negative
effects on row crops. The objective of this study was to investigate the
flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing in corn, soybean and bareground environments
across five locations in the North Central United States.

@@

A Palmer amaranth accession (Kei3) from Perkins County, Nebraska with no
glyphosate resistance was selected for this study (Oliveira et al.,
2021). Three weeks prior to the establishment of each cohort, seeds were
planted in plastic trays containing potting-mix. Emerged seedlings (1
cm) were transplanted into 200 cm\textsuperscript{-3} plastic pots (a
plant pot\textsuperscript{-1}). Palmer amaranth seedlings were supplied
with adequate water and kept under greenhouse conditions at the
University of Wisconsin-Madison, University of Nebraska-Lincoln, and
Western Illinois University; and kept outdoors at the Perkins extension
office in Grant, NE until the 2-3 leaf stage (5 to 8 cm height) when
they were transported to the field.

@@

Fields were conventionally tilled prior to crop planting. Corn and
soybean were planted in 76-cm row spacing (Table 1). Monthly mean air
temperature and total precipitation were obtained using Daymet weather
data from June through September across the five locations in 2018 and
2019 (Correndo et al., 2021) (Figure 1).

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each field
experimental unit was planted with corn, soybean, or maintained as
bareground. The two transplant timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting to assure seedling survival.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant
gender was identified (male or female), plant height was measured from
soil surface to the top of plant. Also, aboveground plant, and all
aboveground plant organs were harvested, then oven dried at 65 C until a
constant weight was reached, and dry biomass (g
plant\textsuperscript{-1}) was recorded. Plants were harvested at
flowering because Palmer amaranth infestations do not occur at the
Wisconsin or Illinois research locations. In our study, all locations
followed the methodology of plant harvest at flowering initiation,
except in Grant, NE. At the Grant, NE, location, all plants from the
first cohort were harvested on July 6, 2018 or July 6, 2019, and all
plants from the second cohort were harvested on August 17, 2018 or July
31, 2019, regardless of flowering status.

@@

Analyses of Palmer amaranth height and biomass were performed with a
linear mixed model using \emph{lmer} function from ``lme4'' package
(Bates et al., 2015). Plant height and biomass were log transformed to
meet model assumption of normality. In the model, crop (bareground,
corn, soybean) and cohort time (first and second) were the fixed effects
and year nested with location were the random effects. Analysis of
variance at \(\alpha\) 0.05 was performed with \emph{anova} function
from ``car'' package (Fox and Weisberg, 2018). Marginal means and
compact letter display were estimated with \emph{emmeans} and \emph{cld}
from packages ``emmeans'' (Lenth et al., 2021) and ``multcomp'' (Hothorn
et al., 2008), respectively.

Palmer amaranth cumulative flowering estimation was determined across
all locations, except Grant, NE. Cumulative flowering estimation was
determined using an asymmetrical three parameter log logistic Weibull
model of the drc package (Ritz et al., 2015):

@@

In this model, \emph{Y} is the cumulative flowering, \emph{d} is the
upper limit (set to 100), and \emph{e} is the inflection point, and
\emph{x} is the day of year (doy).

The doy for 10, 50, and 90\% cumulative flowering were determined using
the \emph{ED} function of drc package. Also, the 10, 50, and 90\% Palmer
amaranth cumulative flowering were compared among crops and cohorts
using the \emph{EDcomp} function of drc package. The EDcomp function
compares the ratio of cumulative flowering using t-statistics, where
P-value \textless{} 0.05 indicates that we fail to reject the null
hypothesis.

A binary logistic regression was fitted to Palmer amaranth gender.
Binary logistic regression is used for predicting binary classes
(Bangdiwala, 2018), such as the probability of a plant being female in a
dioecious species. Prior to the analysis, missing values were removed
from the dataset (including all data from the Grant location). The
resulting dataset was split into 80\% train and 20\% test data. The 80\%
train is used for model training and the 20\% test is used for checking
model performance on unseen dataset. Using the 80\% train data, a
generalized linear model (base R \emph{glm} function) was fitted to
binary response variable, the probability of being female (0 to male and
1 to female). The independent variables were harvest day of year,
height, weight, and crop (without interaction). The model family was
binomial with a logit function. The model fit was assessed through
pseudo R-squared values (McFadden, Cox and Snell, Cragg and Uhler) and
likelihood ratio using \emph{nagelkerke} function from ``rcompanion''
package (Mangiafico, 2021). The marginal effects computation was
performed with Average Marginal Effects (AMEs) at every observed value
of x and averaged across the results (Leeper, 2017) using \emph{margins}
function from ``margins'' package (Leeper et al., 2021). The 20\% test
data was predicted using the \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function from
ROCR package (Sing et al., 2005). The model quality prediction from the
classification algorithm was measured with precision (\emph{precision}
function), recall (\emph{recall} function) and F1-score (\emph{f\_meas}
function) using the ``yardstick'' package (Kuhn et al., 2021). The
precision determines the accuracy of positive predictions (female
plants), recall determines the fraction of positives that were correctly
identified, and F1-score is a weighted harmonic mean of precision and
recall with the best score of 1 and the worst score of 0 (Raoniar,
2021). F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020). The area under the receiver operating curve
(AUC-ROC) was also estimated with performance function using the true
positive and false positive rates. The higher the AUC, better the model
is at distinguishing between female and male Palmer amaranth.

@@

Palmer amaranth plants accumulated more biomass when growing in
bareground compared to plants growing in soybean and corn (figure 3A).
Palmer amaranth plants in the first cohort produced 75.5, 28.3, and 16.3
g plant\textsuperscript{-1} in bareground, soybean and corn,
respectively. Plants from the second cohort produced 62.6 g
plant\textsuperscript{-1} in bareground, followed by 6.3 g
plant\textsuperscript{-1} in soybean, and 1.4 g
plant\textsuperscript{-1} in corn.

Palmer amaranth height was less affected by cohort timing than was
weight, with the exception of plants growing in corn (Figure 3B). Plants
from the first cohort were on average 69.2 cm tall in bareground, which
was not different from the 70.7 cm tall plants from the second cohort
timing (P = 0.74). In addition, no difference in Palmer amaranth height
(69.3 cm) was detected from first cohort plants in soybean to first and
second cohort plants in bareground (P \textgreater{} 0.75). Palmer
amaranth plants from the second cohort were nearly 10 cm shorter
compared to the first cohort in soybeans (P = 0.04). The tallest (85.2
cm, first cohort) and shortest (38.2 cm, second cohort) plants were
observed in corn.

@@

The initiation and duration of Palmer amaranth flowering was strongly
influenced by cohort and surrounding vegetation (Figure 4A, 4B). In the
first cohort, floral initiation (10\% flowering) occurred near the end
of June for all three treatments, at doy 180, 180.9, and 181.7 for
soybean, bareground, and corn, respectively. In the second cohort,
floral initiation occurred earlier on plants growing in the bareground
than plants growing in soybean or corn (doy 203.8 versus doy 210.9 or
216.8, respectively). Palmer amaranth growing in the bareground had the
shortest duration of floral initiation (measured as the difference
between 10\% and 90\% cumulative flowering) in both the first (34 days)
and second (28 days) cohorts. Palmer amaranth growing in soybean had a
shorter duration of flowering than corn in the first cohort (40 days vs
71 days), but a longer duration in the second cohort (50 vs 44 days).

@@

The model goodness of fit was 0.23, 0.32, 0.40 using pseudo R-squared
test from McFadden, Cox and Snell, and Cragg and Uhler, respectively.
The likelihood ratio test showed a p-value of \textless{} 0.00. The
average marginal effects showed that Palmer amaranth growing in corn
resulted in 14.8\% fewer females plants (Table 2). Moreover, increasing
a cumulative flowering unit doy increases the probability of having a
female plant by 0.4\% (Table 2 and Figure5A). A similar trend was
observed for weight (Figure 5B) and height (Figure 5C), where the
probability of being female increased 0.1\% and 0.2\% for each unit
increase of weight (g) and height (cm), respectively (Table 2).

\textbackslash begin\{table\}{[}!h{]}

\textbackslash caption\{\label{tab:unnamed-chunk-3}Average marginal
effects of Palmer amaranth gender logistic model. Factor pararemter
values (crop\_corn and 'crop\_bareground) are calculated relative to
soybean.\} \centering \fontsize{10}{12}\selectfont

\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

weight & 0.001 & 0.000 & 0.000 & 0.001 & 2.179 & 0.029\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} Average Marginal Effects. \textsuperscript{b} Standard Error.}\\
\end{tabular}

\textbackslash end\{table\}

@@

Our study confirmed that Palmer amaranth is well adapted to growing
throughout the Midwestern United States, and is simply limited in its
range by seed dispersal (Davis et al.~2015). We also confirmed Palmer
amaranth's extraordinary plasticity to adapt to different
agroecosystems. For example, Palmer amaranth mimicked crop architecture
in competing for light (Figure 6), where plants growing in corn
allocated resources to height, while plants growing in bareground
allocated resources to numerous branches. Plants growing in the absence
of a crop produced the greatest biomass, responding to a greater
abundance of light, nutrient and water resources (Figure 3A, 3B). These
results support the argument that Palmer amaranth can quickly evolve
life-history traits to adapt to different cultural practices, similar to
observations in a Palmer amaranth response to nitrogen study (Bravo et
al., 2018). Our results highlight Palmer amaranth as a threat to field
crops as breeding more competitive crop varieties is likely to select
more competitive weed biotypes (Bravo et al., 2017).

@@

Floral initiation depends on the complex interaction between a plant's
genetic makeup and environmental conditions (Lang, 1965). We observed a
significant disruption in flowering caused by surrounding vegetation and
cohort timing. Although all three treatments began flowering at the same
time in the first cohort (about doy 180), it took much less time for the
bareground treatment to reach 90\% flowering (34 days versus 40 for
soybean and 71 for corn). The lack of competition (and stress) in the
bareground resulted in plants that flowered earlier, and were they not
harvested, would have had longer reproductive periods, thereby producing
more seed. Similarly, in the second cohort, the window for all plants to
initiate flowering was much shorter for the bareground (28 days)
compared to Palmer amaranth in soybean (50 days) or corn (44 days).
Significantly, plants in the bareground treatment flowered one week
earlier than soybean, and almost two weeks earlier than corn, again
allowing for greater seed production per plant had they not been
harvested at initial flowering. In some cases, early flowering may
confer an evolutionary advantage, provided the plant has an
indeterminate habit and flowering does not restrict plant growth. In an
Iowa study, Palmer amaranth initiated flowering two weeks prior to
waterhemp (Baker, 2021), and this characteristic may partially explain
why Palmer amaranth can displace waterhemp where both occur in the same
agronomic field. In other cases, however, early flower initiation may be
a response to stressful conditions as a plant attempts to reproduce
before running out of resources. In a dioecious species like Palmer
amaranth, exerting stress on plants to manipulate flowering may be
beneficial in limiting seed production (McFarlane et al., 2018;
Schliekelman et al., 2005). For example, when growing under water
stress, there was a seven days flowering mismatch between male and
female plants (Mesgaran et al., 2021), which can minimize plant
outcrossing, reduce seed production and the exchange of resistant
alleles (Jhala et al., 2021). In our study we observed

@@

\leavevmode\hypertarget{ref-kumar2020}{}%
Kumar, V., Liu, R., and Stahlman, P. W. (2020). Differential sensitivity
of {Kansas Palmer} amaranth populations to multiple herbicides.
\emph{Agronomy Journal} 112, 2152--2163.
doi:\href{https://doi.org/10.1002/agj2.20178}{10.1002/agj2.20178}.
